Tempo and mode of promoter evolution as observed in the \textit{Paramecium aurelia} species complex

### 2020-07-30 manuscript.tex
Small changes to phrasing in introduction

A limited number of studies have investigated the evolution of promoters using TSS profiling information. Frith \textit{et al}. \citep{Frith:2006bb} compared TSSs between human and mouse tissues using the first generation of CAGE data. The authors observed ~1250 instances of shifting promoter usage within homologous sites as well as tissue-specific TSS differences between species. Frith and colleagues also found that, consistent with expectations, TSSs with high turnover exhibited less sequence conservation in its promoter region than those without. The authors propose a model of gradual shifting between TSSs via alternative initiation sites, as well as shifting of initiation along the sequence itself. \\
Main et al. \citep{Main:2013dg}, used a separate TSS profiling method in \textit{D. melanogaster} to identify putative orthologous TSSs in three other \textit{Drosophila} taxa: \textit{D. simulans}, \textit{D. pseudoobscura} and \textit{D. sechellia}. The authors found that the location and estimated activities of orthologous TSSs to be generally conserved, but did observe instances of TSS variation that suggest new promoters. Notably, Main \textit{et. al.} found differences in promoter selection in the Alcohol dehydrogenase (ADH) gene across the four species \citep{Main:2013dg}. Working with high-resolution promoter data in human and mouse, Young and colleagues \citep{Young:2015xx} found that more than half (56\$) of tissue-matched promoters were turned over between the two species. Noting that most protein-coding genes in mammals \citep{FANTOMConsortiumandtheRIKENPMIandCLSTDGT:2014hz} have multiple promoters, the authors propose that promoter turnover is an important but overlooked means by which transcriptional modules diverge during evolution. Li and colleagues (2018) investigated the fates of TSSs across  the primate lineage. They found that new TSSs frequently emerge in transposable elements and repetitive regions of DNA, and are more likely to emerge in areas near existing regulatory regions. While this work was very valuable, showing potential routes of regulatory innovation, the study was limited by the fact it only used empirically-defined TSSs in human (as those from other primates were not available), a fact that prevented the study from identifying novel promoters in the other primate taxa in the study. // 

>>>>

A limited number of studies have investigated the evolution of promoters using TSS profiling information. Frith \textit{et al}. \citep{Frith:2006bb} compared TSSs between human and mouse tissues using the first generation of CAGE data. The authors observed ~1250 instances of shifting promoter usage within homologous sites, as well as tissue-specific TSS differences between species. Frith and colleagues also found that, consistent with expectations, TSSs with high turnover exhibited less sequence conservation in its promoter region than those without. The authors propose a model of gradual shifting between TSSs via alternative initiation sites, as well as shifting of initiation along the sequence itself. \\
Main et al. \citep{Main:2013dg}, used a separate TSS profiling method in \textit{D. melanogaster} to identify putative orthologous TSSs in three other \textit{Drosophila} taxa: \textit{D. simulans}, \textit{D. pseudoobscura} and \textit{D. sechellia}. The authors found that the location and estimated activities of orthologous TSSs to be generally conserved, but did observe instances of TSS variation indicative of new promoters. Notably, Main \textit{et. al.} found differences in promoter selection in the Alcohol dehydrogenase (ADH) gene across the four \textit{Drosophila} species \citep{Main:2013dg}. Working with high-resolution promoter data in human and mouse, Young and colleagues \citep{Young:2015xx} found that more than half (56\$) of tissue-matched promoters were lost, or turned over, between the two species. Noting that most protein-coding genes in mammals \citep{FANTOMConsortiumandtheRIKENPMIandCLSTDGT:2014hz} have multiple promoters, the authors propose that promoter turnover is an important but overlooked means by which transcriptional modules diverge during evolution. Li and colleagues (2018) investigated the fates of TSSs across  the primate lineage. They found that new TSSs frequently emerge in transposable elements and repetitive regions of DNA, and are more likely to emerge in areas near existing regulatory regions. While this work was valuable, demonstrating potential routes of regulatory innovation, the study was limited by the fact it only used empirically-defined TSSs in human (as those from other primates were not available), a fact that prevented the study from identifying novel promoters in the other primate taxa in the study. // 
